Short-time simulations of open collective quantum spin systems

### 2019-04-20 soc_squeezing/collective_paper/collective_spin.tex
refactored code in two-time correlator notebook
minor wording change in an appendix

\begin{figure*}
  \centering
  \includegraphics{two_time.pdf}
  \caption{Magnitude and phase of the two-time connected correlator
    $C\p{t}\equiv\abs{C\p{t}}\exp\sp{i\phi\p{t}}$ defined in
    Eq.~\eqref{eq:two_time} for $N=10^4$ spins initially in the
    polarized state $\ket{\X}\propto\p{\ket\up+\ket\dn}^{\otimes N}$
    evolving under the squeezing Hamiltonians in
    Eqs.~\eqref{eq:OAT}--\eqref{eq:TNT}.  Results are shown for time
    evolution without decoherence (leftmost panel), as well as with
    decoherence at rates $\Gamma_\a$ or $\gamma_\a$, respectively
    corresponding to collective or single-spin decoherence through
    jump operators $\hat S_\a$ or $\hat s_\a^{(j)}$.  Collective
    decoherence rates ($\Gamma_\a$) were chosen to be comparable to
    the OAT squeezing strength $\chi$, while single-spin decoherence
    rates ($\gamma_\a$) were scaled by an additional factor of $S=N/2$
    in order to yield comparable effects on the correlator $C\p{t}$ at
    the times shown.  At short times, the correlations captured by
    $C\p{t}$ are essentially insensitive to collective dephasing
    ($\Gamma_\z$), while the magnitude $\abs{C\p{t}}$ is generally
    suppressed by other forms of decoherence.  Whether decoherence
    occurs through spin decay, excitation, or dephasing also leaves a
    signature on the phase $\phi\p{t}$ of $C\p{t}$.  Results computed
    using the TST expansion in Eq.~\eqref{eq:TST} with $M=20$.}
  \label{fig:two_time}
\end{figure*}

@@

In order to compute a quantity such as spin squeezing, we need to
compute correlators of the form $\bk{\S_{\v n}\p{t}}$, where
$\S_{\v n}\p{t}$ is the time-evolved Heisenberg operator corresponding
to $\S_{\v n}$, with $\S_{\v n}\p{0}\equiv\S_{\v n}$.  The order-$M$
truncated short-time (TST) expansion takes
\begin{align}
  \bk{\S_{\v n}\p{t}}
  \to \sum_{k=0}^M \f{t^k}{k!} \sum_{\v m} \bk{\S_{\v m}} T^k_{\v m\v n},
  \label{eq:tutorial_TST}
\end{align}
where $T^k_{\v m\v n}$ are matrix elements of the $k$-th time
derivative operator, given by
\begin{align}
  T^0_{\v m\v n} \equiv
    \begin{cases}
      1 & \v m = \v n, \\
      0 & \t{otherwise}
    \end{cases},
  &&
  T^1_{\v m\v n} \equiv T_{\v m\v n},
  &&
  T^{k>1}_{\v m\v n}
  \equiv \sum_{\v p_1,\v p_2,\cdots,\v p_{k-1}}
  T_{\v m\v p_{k-1}} \cdots T_{\v p_3\v p_2}
  T_{\v p_2\v p_1} T_{\v p_1\v n}.
\end{align}
Matrix elements $T^k_{\v m\v n}$ and initial-time expectation values
$\bk{\S_{\v m}}$ are thus computed on an as-needed basis for any
particular correlator $\bk{\S_{\v n}\p{t}}$ of interest, and combined
according to \eqref{eq:tutorial_TST}.  Note that initial-time
expectation values $\bk{\S_{\v m}}$ are an input to the TST expansion,
and need to be computed separately for any particular initial state of
interest; expectation values with respect to spin-polarized (Gaussian)
states are provided in Appendix \ref{sec:initial_conditions}.  In
practice, we further collect terms in \eqref{eq:tutorial_TST} to write
\begin{align}
  \bk{\S_{\v n}\p{t}} \to \sum_{k=0}^M c_{\v n k} t^k,
  &&
  c_{\v n k} \equiv \f1{k!} \sum_{\v m} \bk{\S_{\v m}} T^k_{\v m\v n},
\end{align}
where $c_{\v n k}$ are time-independent coefficients for the
polynomial time expansion of $\bk{\S_{\v n}\p{t}}$.  After computing
the coefficients $c_{\v n k}$, there is only negligible computational
overhead to compute the correlator $\bk{\S_{\v n}\p{t}}$ for any time
$t$ up to the breakdown time $t_M$ of an order-$M$ TST expansion.

>>>>

\begin{figure*}
  \centering
  \includegraphics{two_time.pdf}
  \caption{Magnitude and phase of the two-time connected correlator
    $C\p{t}\equiv\abs{C\p{t}}\exp\sp{i\phi\p{t}}$ defined in
    Eq.~\eqref{eq:two_time} for $N=10^4$ spins initially in the
    polarized state $\ket{\X}\propto\p{\ket\up+\ket\dn}^{\otimes N}$
    evolving under the squeezing Hamiltonians in
    Eqs.~\eqref{eq:OAT}--\eqref{eq:TNT}.  Results are shown for time
    evolution without decoherence (leftmost panel), as well as with
    decoherence at rates $\Gamma_\a$ or $\gamma_\a$, respectively
    corresponding to collective or single-spin decoherence through
    jump operators $\hat S_\a$ or $\hat s_\a^{(j)}$.  Collective
    decoherence rates ($\Gamma_\a$) were chosen to be comparable to
    the OAT squeezing strength $\chi$, while single-spin decoherence
    rates ($\gamma_\a$) were scaled by an additional factor of $S=N/2$
    in order to yield comparable effects on the correlator $C\p{t}$ at
    the times shown.  At short times, the correlations captured by
    $C\p{t}$ are essentially insensitive to collective dephasing
    ($\Gamma_\z$), while the magnitude $\abs{C\p{t}}$ is generally
    suppressed by other forms of decoherence.  Whether decoherence
    occurs through spin decay, excitation, or dephasing also leaves a
    signature on the phase $\phi\p{t}$ of $C\p{t}$.  Results computed
    using the TST expansion in Eq.~\eqref{eq:TST} with $M=35$.}
  \label{fig:two_time}
\end{figure*}

@@

In order to compute a quantity such as spin squeezing, we need to
compute correlators of the form $\bk{\S_{\v n}\p{t}}$, where
$\S_{\v n}\p{t}$ is the time-evolved Heisenberg operator corresponding
to $\S_{\v n}$, with $\S_{\v n}\p{0}\equiv\S_{\v n}$.  The order-$M$
truncated short-time (TST) expansion takes
\begin{align}
  \bk{\S_{\v n}\p{t}}
  \to \sum_{k=0}^M \f{t^k}{k!} \sum_{\v m} \bk{\S_{\v m}} T^k_{\v m\v n},
  \label{eq:tutorial_TST}
\end{align}
where $T^k_{\v m\v n}$ are matrix elements of the $k$-th time
derivative operator, given by
\begin{align}
  T^0_{\v m\v n} \equiv
    \begin{cases}
      1 & \v m = \v n, \\
      0 & \t{otherwise}
    \end{cases},
  &&
  T^1_{\v m\v n} \equiv T_{\v m\v n},
  &&
  T^{k>1}_{\v m\v n}
  \equiv \sum_{\v p_1,\v p_2,\cdots,\v p_{k-1}}
  T_{\v m\v p_{k-1}} \cdots T_{\v p_3\v p_2}
  T_{\v p_2\v p_1} T_{\v p_1\v n}.
\end{align}
Matrix elements $T^k_{\v m\v n}$ and initial-time expectation values
$\bk{\S_{\v m}}$ are thus computed on an as-needed basis for any
particular correlator $\bk{\S_{\v n}\p{t}}$ of interest, and combined
according to \eqref{eq:tutorial_TST}.  Note that initial-time
expectation values $\bk{\S_{\v m}}$ are an input to the TST expansion,
and need to be computed separately for any given initial state;
expectation values with respect to spin-polarized (Gaussian) states
are provided in Appendix \ref{sec:initial_conditions}.  In practice,
we further collect terms in \eqref{eq:tutorial_TST} to write
\begin{align}
  \bk{\S_{\v n}\p{t}} \to \sum_{k=0}^M c_{\v n k} t^k,
  &&
  c_{\v n k} \equiv \f1{k!} \sum_{\v m} \bk{\S_{\v m}} T^k_{\v m\v n},
\end{align}
where $c_{\v n k}$ are time-independent coefficients for the
polynomial time expansion of $\bk{\S_{\v n}\p{t}}$.  After computing
the coefficients $c_{\v n k}$, there is only negligible computational
overhead to compute the correlator $\bk{\S_{\v n}\p{t}}$ for any time
$t$ up to the breakdown time $t_M$ of an order-$M$ TST expansion.
